Step by Step: Using Step Selection Analysis to Simulate Dispersal and Assess Landscape Connectivity

### 2021-03-25 Chapter_1/04_Manuscript/Manuscript.tex
Improved rasterization speed drastically

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}

During the past two decades, least-cost analysis and circuit theory have been
the two workhorses in connectivity analyses. This is largely thanks to their
ease of use and intuitive nature which has facilitated the application of the
methods across a broad range of animals. However, recent innovations in movement
ecology have brought forward new methods that are similarly intuitive but
provide several advantages over traditional connectivity modelling techniques.

Here, we propose the simulation of dispersal trajectories as a much more generic
way to identify dispersal barriers and potential movement corridors. Moreover,
we show how such simulations can be used to identify zones of high
human-wildlife conflicht, as well as to pinpoint potential "dispersal traps"
where dispersing individuals get trapped due to unavailability of suitable
alternatives. Finally, we discuss the benefits and pitfalls of dispersal
simualtions.
\end{abstract}

@@

\section{Introduction}
% Dispersal & Connectivity
Dispersal of individuals is an important process governing the dynamics wild
animal populations that are distributed in space \citep{Clobert.2012}. It is
defined as the movement of individuals from their natal location to the site of
first reproduction \cite{Howard.1960} and allows to avoid inbreeding
\citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012}, to rescue small and
unviable populations \citep{Brown.1977}, and to promote the colonization of
unoccupied habitats \citep{Hanski.1998, MacArthur.2001}. Information on movement
behavior during dispersal and knowledge about the factors that limit dispersal
is therefore critical for a comprehensive understanding of landscape
connectivity and population viability \citep{Baguette.2013, Vasudev.2015}.

% Technological Advancements
Thanks to novel technologies developed over the past decades, particularly of
GPS/Satellite radio-collars, the study of dispersal has accelerated
\citep{Jonsson.2016, Williams.2019}. Additionally, the advent of publicly
accessible satellite imagery and sophisticated remote sensing techniques to
represent the physical landscape through which individuals disperse has pushed
the field towards a  ``golden age of animal tracking'' \citep{Kays.2015}.
Concurrently, the increased availability of large amounts of empirical data and
an increased computational power have led to the development of numerous
analytical methods to study movement behavior during dispersal
\citep{Boyce.2002, Fortin.2005, Cushman.2010, Zeller.2012, Elliot.2014,
Diniz.2019}.

% Measuring Connectivity
Resource selection functions \citep{Boyce.2002} and derived methods such as step
selection functions \citep{Fortin.2005} and path selection functions
\citep{Cushman.2010} have proven particularly useful for studying animal
movement and landscape connectivity. These methods allow estimating habitat
preferences of the focal species by comparing covariates at locations visited by
the animal to the scame covariates at locations available to, but not visited by
the animal. The so estimated preferences can then be used to predict a
permeability surface, indicating the expected ease at which an animal can
traverse a given area \citep{Zeller.2012}. Utimately, the permeability surface
serves as foundation to a connectivity model and to reveal movement corridors.
In this regard, the literature has long relied on least-cost methods
\citep{Adriaensen.2003} and circuit theory \citep{McRae.2006, McRae.2008}.
Unfortunately, both of these approaches make several limiting assumptions that
are hardly ever met in reality, particularly by dispersing animals. For
instance, to calculate a least-cost path one has to pre-define start- and
end-points. This implicitly assumes that animals have a preconceived end-point
in mind and choose a cost-minimizing route accordingly. To calculate a
least-costly route, one also needs to assume that animals have an infinite
perceptual range and that a route can always be realized, regardless of the
imposed costs. In contrast to least-cost methods, circuit theory assumes a
perceptual range of a single pixel, which is equally unrealistic and rarely
captures the scale of selection (cite someone). Moreover, directional biases
cannot be captured by circuit theory as a random walk is implicitly assumed. In
reality, however, dispersers often depict directional persistence and move in an
almost straight line to cover much ground very quickly. A mix of the two methods
can be seein in least-cost corridor approaches or randomized shortest paths have
been proposed to mitigate these issues by allowing animals to deviate from the
``optimal'' least-costly route \citep{Pinto.2009, Panzacchi.2016,
VanMoorter.2021}, they still fail to render a more realistic perceptual range of
the animal. In addition, neither of the two methods render the temporal
dimension of movement. Thusfar, only few have explored the usefulness of
individual based movement models when estimating landscape connectivity
\citep{Kanangaraj.2013, Hauenstein.2019, Zeller.2020}.

Nevertheless, collecting data on dispersers remains logistically challenging due
to dispersers' inherent mobility and due to the fact that dispersers only make
up a small fraction of the entire population, thereby making it difficult for
scientists to target apporpriate individuals for their study.

An animals movement trajectory can be seen as the result of an interplay between
habitat and movement preferences.

In recent decades, least-cost analysis has become the workhorse to study
landscape connectivty at large scales. Originally introduced by XX, the concept
and relating methods have been refined to more realistically render animal's
movement capabilities and to determine valuable movement corridors that require
protection. (talk a bit about different methods of LCPs, see methods in
gdistance package). More recently, more attention has been assigned to the
collection of movement data is better tailored towards assessing landscape
connectivity. In particular, it has been proposed and verified that relocation
data collected dispersing individuals leads to more accurate and reliable
estimates of landscape connectivity. This superiority of dispersal data in
comparison to data that stem from residents is mainly owed to the fact that
animals behave vastly different during residency. Despite substantial advances
in ecologists ability to track animals in space and time, dispersal remains one
of the most difficult behavioral modes to observe in wild animals. Especially
for wide ranging and long-lived species, dispersal is difficult to predict and
observe, such that data remains scarce.

Many connectivity modelling techniques, especially least-cost analysis,
implicitly assume that the studied animal has partial or even complete knowledge
of the landscape and associated movement costs. While this assumption may be
reasonable for migrating animals that move between a limited number of habitats,
yet it is unlikely to hold for dispersers. In contrast to migrants, dispersers
move into unknown territory and are therefore confronted with novel landscapes.
Consequently, dispersers are more likely to adjust their movement behavior
\textit{ad hoc} instead of preplanning an entire trajectory. As a result,
methods that assume complete knowledge of the landscape and subsequent optimal
movement behavior likely misrepresent true dispersal behavior. As such, methods
that allow a more random approach to a dispersers movement behavior may more
realistically render the movement corridors of dispersing individuals.

However, these models can not account for a correlation between turning angles
and step lengths, unless the two are sampled jointly from a copula distribution
(Hodel et al.).

A relatively recent addition to the step selection family has been integrated
step selection analysis.... Thanks to recently published papers and packages,
the suite of step selection functions has become accessible to most researchers
working with relocation data. Despite this, the use of such methods rarely
extends beyond assessing relative habitat preferences. However, once
parametrized, integrated step selection models could potentially serve as
powerful framework to simulate animal movement and assess landscape
connectivity. Here, we exemplify such analyses and show how ... coud be applied
to investigate ...

In contrast to Signer et et al, our focus lied on the transient utilisation of
the landscape.

Reliable identification of dispersal corridors will become increasingly
important with the uprise of ever-growing and often transboundary conservation
areas. One such instance is the KAZA-TFCA, a massive conservation area spanning
five countries and over 520'000 km\textsuperscript{2}. The KAZA holds the
potential of re-establishing dispersal routes for many of its protected species,
including the african wild dog \textit{Lycaon pictus}.  Persecution by humans,
habitat loss, and reduced connectivity are major causes of the decline of the
species \citep{Woodroffe.2012}. In result, the species currently marks the
KAZA's most endangered large carnivore and has been a assigned a very high
conservation priority. Importantly, due to their inherent mobility and intrinsic
need for vast undisturbed landscapes, AWDs have been proposed as surrogate
species for landscape connectivity (see recent paper on multispecies
connectivity). Nevertheless, the species has received little attention in the
connectivity literature, mainly due to the difficulty in observing wild dog
dispersal. In a previous paper, we addressed this issue and developed a habitat
selection model based on which we predicted landscape connectivity using
least-cost corridor analysis. We now expand on this knowledge and develop a more
detailed movement model of dispersing wild dogs. We then use this model to
simulate thousands of dispersers moving throughout the KAZA. Based on said
simulations, we compute heatmaps and identify potential dispersal hotspots and
compare them to the dispersal routes identified in (Hofmann ....). We also apply
the data to feed a network-analysis, based on which we determine network-metrics
pertinent to landscape connectivity.

@@

\subsection{GPS Relocation Data}
Between 2011 and 2020, we collected GPS relocation data on dispersing wild dogs
from a free-ranging wild dog population inhabiting the Okavango Delta in
northern Botswana. We identified potential dispersers based on age, number of
same‐sex siblings, pack size, and presence of unrelated individuals of the
opposite sex in their pack \citep{McNutt.1996, Behr.2020}. We immobilized
selected individuals using a cocktail of Ketamine/Xylazine/Atropine
\citep{Osofsky.1996, Cozzi.2020} which we injected with a dart, fired from a
CO\textsubscript{2}-pressurized gun (\textit{DAN-Inject, Denmark}). The
corresponding protocols have been approved by the Ministry of Environment,
Natural Resources Conservation and Tourism of Botswana (permit EWT 8/36/4
XXXVI). After immobilization, individuals were fitted with GPS/Satellite radio
collars (\textit{Vertex Lite; Vectronic Aerospace GmbH, Berlin}) that included
an automated drop-off mechanism. Handling and collaring of all individuals was
carried out and supervised by a Botswana-registered wildlife veterinarian. All
of the immobilized individuals quickly rejoined their pack after the procedure.
From all collared individuals, 16 individuals eventually dispersed in separate
same-sex coalitions and their trajectories were successfully recorded (7 female
and 9 male coalitions).

During dispersal, collars were programmed to record a GPS fix every 4 hours.
Recorded relocations were regularly transmitted over the Iridium satellite
system, which allowed remote tracking of individuals, even if they left the main
study area and crossed international borders. To effectively distinguish between
periods of residency and dispersal, we applied the net-squared displacement
metric to the observed movement data. This metric measures the squared Euclidean
distance of a collared individual to a reference point \citep{Borger.2012},
which we set at the center of each disperser's natal home range. Hence,
dispersal was deemed to have started when an individual left its natal home
range and ended when the individual became stationary again. For the purpose of
this study, we discarded andy data that was collected during residency. Because
previous research could not find any differences between males and females
during dispersal \citep{Woodroffe.2019, Cozzi.2020}, we did not distinguish
sexes for the sake of our analyses. Ultimately, we converted the collected GPS
relocations to steps, where a step represented the straight-line distance
travelled between two consequtive GPS relocations \citep{Turchin.1998}.

@@

\subsection{Movement Model}
We used integrated step selection functions (iSSF) to parametrize a mechanistic
movement model of dispersing wild dogs. In this framework, observed steps are
contrasted to random steps that the animal could have realized, but decided not
to. To compare steps, it is assumed that animals assign a relative selection
score \(w(x)\) to each step. The score depends on the covariates \(X\)
experienced along a step, as well as the animals preferences towards these
covariates \(\beta\). The probability of a step being realized is then
contingent on the step's selection score, as well as the selection scores of all
other step in the same stratum. In contrast to regular step selection analysis,
\textit{integrated} step selection analysis allows simulatneous inference on
movement and habitat preferences of the studied animal. Moreover, potential
interactions between habitat and movement preferences can be included to
investigate how movement behavior depends on habitat characteristics. In result,
the method produces less biased selection estimates and allows to apply the
resulting model as a fully mechanistic movement model based on which movement
can be simulated. To conduct iSSF analysis, we paired each observed step with 24
random steps. Random steps basically resembled potential alternatives that the
animal could have realized but decided not to. We generated random steps by
sampling random turning angles from a uniform distribution (\(-\pi, +\pi\)) and
step lengths from a gamma distirbution that was fitted to observed steps (scale
= 6'308, shape = 0.37). While the number of random steps is inversely
proportional to the sampling error \citep{Avgar.2016}, we found only minor
differences in model performance when sampling additional random steps and
deemed 24 random steps sufficient. Along each step, we extracted spatial
covariates using the \textit{velox} package and calculated the movement metrics
\(sl\), \(log(sl\)), and \(cos(ta)\). To facilitate model convergence, we
standardized all continuous covariates to a mean of zero and a standard
deviation of one. We then used the r-package \textit{glmmTMB} to fit mixed
effects conditional logistic regression models as proposed by \citep{Muff.2020}.

@@

\subsection{Dispersal Simulation}
We used the most parsimonious movement model to simulate dispersing wild dogs
departing from the above described source points. The simulation resembled an
inverted iSSF and was set up as follows. First, we determined a source point at
which a disperser was initiated. To allow calculation of turning angles, we
assumed a random orientation of the animal. Second, we proposed a set of 25
random steps leaving from the source point. We generated random steps by
sampling turning angles from a uniform distribution (\(-\pi, +\pi\)) and step
lengths from our fitted gamma distribution. To prevent unrealistically large
steps, we capped sampled step lengths at 35km, which is the farthest distance
travelled by our dispersers within 4 hours. Third, along each random step we
extracted habitat covariates and calculated movement covariates. Fourth, we
applied the parametrized movement model to predict the selection scores \(w(x)\)
for all steps. Fifth, we translated selection scores into probabilitities by
applying equation xy. Sixth, we sampled one of the proposed based on assigned
probabilities and calculated the animal's new position. We then repeated steps
two to six until a total of 2000 steps were realized.

To mitigate the influence of edge effects, we added a buffer zone around each
covariate layer. The values within the buffer zone where resampled from the true
data. As such, simulated dispersers were allowed to leave and reenter the study
area through this buffer zone. In case one of the proposed random steps left the
outer boundary of the buffer, we replaced the step with an alternative step that
lied fully within the buffer zone. Thus, we force the disperser to bounce of the
borders of the study area.

@@

\subsection{Betweenness}

\subsection{Network Analysis I: Areas Reached}
Based on the simulated trajectories we identified to which other areas each
source area is connected. For instance, if a trajectory originated at source
point one and intersected with source areas two and three, we assumed that
source areas two and three were within reach of source are one. To get a sense
of the strength of connections, we also calculated how often each of these
connections was realized. This procedure resulted in 6 x 2 edge lists which we
further used to generate weighted networks and to calculate network metrics.

\subsection{Network Analysis II: Betweenness}
We coerced all simulated trajectories into a network consisting of vertices
(relocations) and edges (connections between relocations). To do so, we created
raster layers at multiple resolutions and identified each trajectory's
transition matrix on these rasters see figure xx \citep{BastilleRousseau.2018}.
We then merged the transition matrices of all trajectories and calculated
cumulative transitions between all raster-cells. This resulted in an edge-list,
containing all observed from-to connections as well as their frequency. Using
this edge-list, we generated a weighted graph using the r-package
\textit{igraph}. Based on this graph we calculated betweenness scores as well as
the degree of each raster cell. Betweenness indicates how often a specific
raster-cell lies on a shortest path between two other raster-cells and is a
useful metric to detect movement corridors. Degree, on the other hand, indicates
how many connections a raster cell has and therefore serves to illustrate the
xxx of specific nodes. When calculating betweenness, we used the transition
frequency as weighting factor. That is, a higher transition frequency
contributed to a higher betweenness score.

\subsection{Network Analysis III: Transitions}

>>>>

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
For many species, a sufficient degree of landscape connectivity is a crucial
prerequisite for long-term species persistence. This is especially true for
animals that exhibit long-distance dispersal, as dispersers typically cross a
variety of habitats and potentially move into unfamiliar territory. Estimating
habitat permeability and quantifying landscape connectivity have therefore
become two major tasks in conservation ecology. For this, least-cost analysis
and circuit theory have been the two workhorses during the past two decades.
This is largely owed to their ease of use and intuitive nature which has
facilitated the application of the methods across a broad range of the animal
kingdom. However, both methods make several restricting assumptions that limit
their usefulness in reality.

Recent innovations in movement ecology, particularly on step selection analysis,
have brought forward novel ways to look at connectivity. For instance,
integrated step selection functions provide a means to parametrize a fully
mechanistic movement model based on which virtual dispersers could be simulated.
While this approach has been used to infer habitat utilization, it may also
serve to highlight landscape connectivity and pinpoint dispersal corridors.

Here, we propose the simulation of dispersal trajectories as a much more generic
way to identify dispersal barriers and potential movement corridors. To achieve
this, we applied integrated step selection analysis and uses data from 16
dispersing wild dogs to parametrize a fully mechanistic movement model for
dispersing African wild dogs. The model rendered dispersers' habitat and
movement preferences, as well as potential interactions between them. Based on
the model, we simulated myriads of dispersal trajectories and inferred dispersal
corridors across the earth's largest transboundary conservation area, the
Kavango-Zambezi Transfrontier Conservation Area. We also exemplify how such
simulations could be analysed using network theory. Finally, we discuss the
benefits and pitfalls of dispersal simualtions and highlight potential
improvements to be made in the future.
\end{abstract}

@@

\section{Introduction}
% Importance of Dispersal & Connectivity
Dispersal of individuals is an important process governing the dynamics wild
animal populations that are distributed in space \citep{Clobert.2012}. It is
defined as the movement of individuals from their natal location to the site of
first reproduction \cite{Howard.1960} and allows species to avoid inbreeding
\citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012}, to rescue small and
unviable populations \citep{Brown.1977}, and to promote the colonization of
unoccupied habitats \citep{Hanski.1998, MacArthur.2001}. Successfull dispersal
requires a sufficient degree of landscape connectivity, which is why the
protection of dispersal corridors has become a major goal in conservation
science. Information on movement behavior during dispersal and knowledge about
the factors that limit dispersal is therefore critical for a comprehensive
understanding of landscape connectivity and population viability
\citep{Baguette.2013, Vasudev.2015}. In addition, reliable modelling techniques
to identify dispersal corridors based on empirical data are necessary
\citep{Diniz.2019}.

% Advancements in GPS Technology and Movement Ecology
Thanks to novel technologies developed over the past decades, particularly of
GPS/Satellite radio-collars, the study of dispersal and connectivity using
telemetry data has accelerated \citep{Jonsson.2016, Williams.2019}.
Additionally, the advent of publicly accessible satellite imagery and
sophisticated remote sensing techniques to represent the physical landscape
through which individuals disperse have heralded the ``golden age of animal
tracking'' \citep{Kays.2015}. Concurrently, the increased availability of large
amounts of empirical data and an increased computational power have led to the
development of several modelling techniques that allow studying movement
behavior during dispersal and highlighting major movement corridors
\citep{Boyce.2002, Fortin.2005, Cushman.2010, Zeller.2012, Diniz.2019}.

% Resource Selection & Connectivity
\textit{Resource selection functions} \citep{Boyce.2002} and derived methods
such as \textit{step selection functions} \citep{Fortin.2005} and \textit{path
selection functions} \citep{Cushman.2010} have proven particularly useful for
studying animal movement \citep{Fieberg.2020}. These methods allow estimating
habitat preferences of the focal species by comparing covariates at locations
visited by the animal to the scame covariates at locations available to, but not
visited by the animal. The so estimated preferences can then be used to predict
a permeability surface, indicating the expected ease at which an animal can
traverse a given area \citep{Zeller.2012}. Utimately, the permeability surface
serves as input to a connectivity model that is used to reveal movement
corridors. In this regard, two of the most prominent connectivity models in the
literature are least-cost path analysis \citep{Adriaensen.2003} and circuit
theory \citep{McRae.2006, McRae.2008}.

% Least-cost path
In least-cost path analysis, connectivity is estimated by identifying
least-costly routes between pre-defined start- and endpoints
\citep{Adriaensen.2003}. For this, the permeability surface is converted into a
network graph and permeability scores are translated into probabilities for
moving from one cell to another. Least-costly routes between start- and
endpoints can then be computed using Dijkstra’s algorithm \citep{Dijkstra.1959}.
Although intuitively appealing, the method suffers from several weaknesses.
First of all, a least-cost path always exist, even if the associated costs are
unreasonably high. Second, it assumes that animals have a preconceived end-point
in mind and choose a cost-minimizing route accordingly. While this assumption
may be reasonable for migrating animals that move between a limited number of
habitats, it is unlikely to hold for dispersers that move over long distances
into unknown territory \citep{Koen.2014, Abrahms.2017, Cozzi.2020}. Finally, the
method also requires that animals have an infinite perceptual range, otherwise
they could not compute an optimal path. Some of these issues have been addressed
using alternative least-cost approaches. For instance, instead of computing a
least-cost path, a least cost-corridor that also considers slightly suboptimal
routes \citep{Pinto.2009} can be calculated. The corridor relaxes the assumption
of perfect knowledge and overcomes the single pixel width issue. Alternatively,
the randomized least-cost path has been proposed, which allows animals to
deviate from the least-cost route by a pre-defined factor \(\theta\)
\citep{Panzacchi.2016, VanMoorter.2021}.

% Circuit theory
In contrast to least-cost approaches, circuit theory makes use of electrical
circuit theory to quantify landscape connectivity \citep{McRae.2006,
McRae.2008}. Here, the permeability surface is converted into an electrical
circuit containing nodes and resistor, where resistance values depend on habitat
permeability. A source point is connected to a current source, whereas an
endpoint is grounded. Hence, the current flowing through each node in the
circuit can be computed and is supposed to resemble the relative frequency at
which each location in the study area is being visited. By revealing not only a
single path or corridor, circuit theory overcomes the single-pixel width issue
inherent to least-cost approaches. Intrinsically, circuit theory is based on the
assumption that animals follow a random walk and can therefore not render
directional biases. In reality, however, dispersers often move in a very
directional fashion as they want to cover as much ground in as little time as
possible. Furthermore, it is assumed that individuals have a perceptual range of
a single pixel, which rarely captures the true spatial scale of selection (cite
someone).

% Temporal scale
Neither least-cost analyes or circuit theory are capable of rendering the
temporal dimension of movement. Such influences could, however, be studied using
individual based movement models. Still, only few have explored these
possibilities \citep{Kanangaraj.2013, Hauenstein.2019, Zeller.2020}.

% Step selection analysis
While step selection functions were initially intended to learn about habitat
preferences of the focal species \citep{Fortin.2005}, recent improvements may
elevate its applicability into other areas. Originally, \cite{Fortin.2005}
proposed step selection functions to account for serial autocorrelation inherent
to GPS relocation data. In this method, covariates along realized steps (the
connecting lines between two consecutive relocations; ; \citealp{Turchin.1998})
are contrasted to covariates along  ``alternative''  or  ``random steps'' that
are generated by randomly sampling turning angles and step lengths based on
observed movement characteristics. It is then assumed that animals assign a
selection score \(w(x)\) of the following form to each step:

\begin{equation}
\label{EQ1}
  w(x) = exp(\beta_1 x_1 + \beta_2 x_2 + ... + \beta_n x_n)
\end{equation}

The selection score \(w(x)\) of each step depends on its associated covariates
(\(x_1, x_2, ..., x_n\)) and on the animal's relative selection strengths (i.e.
preferences) towards these covariates (\(\beta_1, \beta_2, ..., \beta_n\)). The
probability that a step \(i\) is realized \(P(Y_{i} = 1\)) then depends on the
step's selection score, as well as on the selection scores of all alternative
steps:

\begin{equation}
\label{EQ2}
  P(Y_{i} = 1 | Y_{1} + Y_{2} + ... + Y_{i} = 1) =
  \frac{w(x_{i})}{w(x_{1}) + w(x_{2}) + ... + w(x_{i})}
\end{equation}

\noindent Preferences of interest, i.e. the \(\beta\)'s, can be estimated by
comparing realized (scored 1) and random (scored 0) steps in a conditional
logistic regression model \citep{Fortin.2005}. In this model, positive
\(\beta\)-coefficients indicate selection of a covariate, whereas negative
\(\beta\)-coefficients indicate avoidance of a covariate. To deal with multiple
individuals, one can either average estimates of individual models
\citep{Murtaugh.2007, Fieberg.2010} or apply mixed effects conditional logistic
regression analysis as recently proposed by \cite{Muff.2020}.

% Integrated step selection analysis
An animals movement trajectory can be seen as the result of an interplay between
habitat and movement preferences. Traditional step selection functions cannot
account for a correlation between turning angles and step lengths, unless the
two are sampled jointly from a copula distribution \citep{Hodel.2021}. Recently,
however, SSFs have been generalized to \textit{integrated} SSFs (iSSFs), which
tear apart movement and habitat preferences of the studied animals
\citep{Avgar.2016}. The method thus allows to render an animals preferences with
respect to prevailing habitat conditions, as well as with respect to movement
characteristics. This reduces potential biases in estimated preferences and
enables parametrization of a fully mechanistic movement model from which
movement and space use can be simulated \citep{Avgar.2016, Signer.2017}.
\cite{Signer.2017} used integrated step selection analysis to simulate steady
state utilization distributions of resident animals. However, the degree to
which such simulations could be used to simulate dispersers and infer movement
corridors is unknown.

% Study Species & Study Area
One of the species for which long-term species persistence relies on sufficient
landscape connectivity is the endangered African wild dog \textit{Lycon pictus}.
This species has once been wide-spread across sub-Saharan Africa but disappeared
from a vast majority of its historic range due to ongoing persecution by humans,
habitat destruction, and deadly diseases. As of today, only 6'000 free-ranging
individuals remain in small and spatially scattered subpopulations. Within those
subpopulations, wild dogs form cohesive packs comprising 8 to 12 adults and
their offspring \cite{McNutt.1995}. After reaching sexual maturity, male and
female offspring form same-sex coalitions and disperse from their natal pack
\citep{McNutt.1996, Behr.2020}. New packs are formed when dispersing coalitions
join unrelated opposite-sex dispersing coalitions \citep{McNutt.1996}.
Dispersing wild dogs can cover several hundred kilometers across a variety of
landscapes \citep{DaviesMostert.2012, Masenga.2016, Cozzi.2020, Hofmann.2021}.
One of the few strongholds for this species lies near the Moremi Game Reserve in
northern Botswana, which is part of the world's largest transboundary protected
area, namely the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA).
This area has originally been intended to facilitate migration of elephants, but
is expected to benefit a multitude of other species \citep{Elliot.2014,
Brennan.2020, Hofmann.2021}.

In a previous paper, we assessed landscape connectivity within the KAZA-TFCA for
dispersing African wild dogs using a least-cost corridor approach. For this, we
fitted a basic habitat selection model based on which we predicted landscape
permeability. We now expand on this knowledge and develop a more detailed
movement model of dispersing wild dogs. We then use this model to simulate
thousands of dispersers moving throughout the KAZA. Based on said simulations,
we compute heatmaps and identify potential dispersal hotspots and compare them
to the dispersal routes identified in \citep{Hofmann.2021}. We also showcase how
simulated dispersal data can be analysed using network-analysis and how network
metrics relevant to landscape connectivity can be computed. Our results suggest
that a simulation based approach yields several major benefits over traditional
connectivity modelling techniques. Most importantly, simulations provide a more
generic view on how connectivity emerges and to which degree connectivity
depends on the dispersal duration. In addtion, by generating proper dispersal
trajectories, network theory can be applied to calculate network metrics that
are pertinent to connectivity analysis. Finally, we put forward additional
opportunities using simulations that go beyond the scope of this paper.

Reliable identification of dispersal corridors will become increasingly
important with the uprise of ever-growing and often transboundary conservation
areas. One such instance is the KAZA-TFCA, a massive conservation area spanning
five countries and over 520'000 km\textsuperscript{2}. The KAZA holds the
potential of re-establishing dispersal routes for many of its protected species,
including the african wild dog \textit{Lycaon pictus}.  Persecution by humans,
habitat loss, and reduced connectivity are major causes of the decline of the
species \citep{Woodroffe.2012}. In result, the species currently marks the
KAZA's most endangered large carnivore and has been a assigned a very high
conservation priority. Importantly, due to their inherent mobility and intrinsic
need for vast undisturbed landscapes, AWDs have been proposed as surrogate
species for landscape connectivity (see recent paper on multispecies
connectivity). Nevertheless, the species has received little attention in the
connectivity literature, mainly due to the difficulty in observing wild dog
dispersal.

@@

\subsection{GPS Relocation Data}
Between 2011 and 2020, we collected GPS relocation data on dispersing wild dogs
from a free-ranging wild dog population inhabiting the Okavango Delta in
northern Botswana. We identified potential dispersers based on age, number of
same‐sex siblings, pack size, and presence of unrelated opposite-sex individuals
in the pack \citep{McNutt.1996, Behr.2020}. We immobilized individuals using a
cocktail of Ketamine/Xylazine/Atropine \citep{Osofsky.1996, Cozzi.2020},
injected with a dart that was fired from a CO\textsubscript{2}-pressurized gun
(\textit{DAN-Inject, Denmark}). Protocols for the anesthization protocols were
approved by the Ministry of Environment, Natural Resources Conservation and
Tourism of Botswana (permit EWT 8/36/4 XXXVI). After immobilization, individuals
were fitted with GPS/Satellite radio collars (\textit{Vertex Lite; Vectronic
Aerospace GmbH, Berlin}) that included an automated drop-off mechanism. Handling
and collaring of all individuals was carried out and supervised by a
Botswana-registered wildlife veterinarian. All of the immobilized individuals
quickly rejoined their pack within one hour after the procedure. Out of all
collared individuals, 16 eventually dispersed in separate same-sex coalitions
and their trajectories were successfully recorded (7 female and 9 male
coalitions).

During dispersal, collars were programmed to record a GPS fix every 4 hours.
Recorded relocations were regularly transmitted over the Iridium satellite
system, which allowed remote tracking of individuals, even if they left the main
study area and crossed international borders. To effectively distinguish between
periods of residency and dispersal, we applied the net-squared displacement
metric to the observed movement data. This metric measures the squared Euclidean
distance of a collared individual to a reference point \citep{Borger.2012},
which we set at the center of each disperser's natal home range. Hence,
dispersal was deemed to have started when an individual left its natal home
range and ended when the individual became stationary again. For the purpose of
this study, we discarded andy data that was collected during residency. Because
previous research could not find any differences between males and females
during dispersal \citep{Woodroffe.2019, Cozzi.2020}, we did not distinguish
sexes for the sake of our analyses. Ultimately, we converted the collected GPS
relocations to steps, where each step represented the straight-line distance
travelled between two consequtive GPS relocations \citep{Turchin.1998}.

@@

\subsection{Movement Model}
We used integrated step selection functions (iSSF) to parametrize a mechanistic
movement model of dispersing wild dogs. In this framework, observed steps are
contrasted to random steps that the animal could have realized, but decided not
to. To compare steps, it is assumed that animals assign a relative selection
score \(w(x)\) to each step. The score depends on the covariates \(X\)
experienced along a step, as well as the animals preferences \(\beta\) towards
these covariates. The probability of a step being realized is then contingent on
the step's selection score, as well as the selection scores of all other step in
the same stratum. In contrast to regular step selection analysis,
\textit{integrated} step selection analysis allows simulatneous inference on
movement and habitat preferences of the studied animal. Moreover, potential
interactions between habitat and movement preferences can be included to
investigate how movement behavior depends on habitat characteristics. In result,
the method produces less biased selection estimates and allows to apply the
resulting model as a fully mechanistic movement model based on which movement
can be simulated. To conduct iSSF analysis, we paired each observed step with 24
random steps. Random steps basically resembled potential alternatives that the
animal could have realized but decided not to. We generated random steps by
sampling random turning angles from a uniform distribution (\(-\pi, +\pi\)) and
step lengths from a gamma distirbution that was fitted to observed steps (scale
= 6'308, shape = 0.37). While the number of random steps is inversely
proportional to the sampling error \citep{Avgar.2016}, we found only minor
differences in model performance when sampling additional random steps and
deemed 24 random steps sufficient. Along each step, we extracted spatial
covariates using the \textit{velox} package and calculated the movement metrics
\(sl\), \(log(sl\)), and \(cos(ta)\). To facilitate model convergence, we
standardized all continuous covariates to a mean of zero and a standard
deviation of one. We then used the r-package \textit{glmmTMB} to fit mixed
effects conditional logistic regression models as proposed by \citep{Muff.2020}.

@@

\subsection{Dispersal Simulation}
We used the most parsimonious movement model to simulate a total 80'000
dispersing wild dogs across the KAZA-TFCA. The simulation basically resembled an
inverted iSSF and was set up as follows. First, we sampled a source point at
which a disperser was initiated and assumed a random initial orientation of the
animal. The orientation was essential to permit calculation of relative turning
angles. Second, we generated a set of 25 random steps by sampling turning angles
from a uniform distribution (\(-\pi, +\pi\)) and step lengths from the fitted
gamma distribution. Similarly to our input data, each step corresponded to a
temporal difference of 4 hours. To prevent unrealistically large steps, we
capped sampled step lengths at 35km, which corresponds to the farthest distance
travelled by one of our dispersers within 4 hours. Third, along each random step
we extracted habitat covariates and calculated movement covariates. Fourth, we
applied the parametrized movement model to predict the selection scores \(w(x)\)
for all steps. Fifth, we translated selection scores into probabilities by
applying \Cref{EQ2}. Sixth, we sampled one of the random steps based on assigned
probabilities and determined the animal's new position. We then repeated steps
two to six until a total of 2'000 steps were realized.

Inevitably, some simulated dispersers would approach a map boundary and one of
the generated random steps would leave the study area. However, beyond the study
area no covariates could be extracted and no selection scores can be calculated.
To mitigate the influence of such edge effects, we buffered all covariate layers
by xx percent and filled the buffer zone with covariate values resampled from
the observed data. As such, simulated dispersers were allowed to leave and
reenter the study area through this buffer zone. In case one of the proposed
random steps left the buffer, we resampled the corresponding step until it lied
fully within the buffer zone. This implies that we forced the disperser to
bounce off the borders of the buffer zone.

To meaningfully select source points, we randomly sampled coordinates within
protected areas larger (\(>\) 700 km\textsuperscript{2}), which conforms to the
average home range requirement of resident wild dogs \citep{Pomilia.2015}. By
distributing source points randomly, the number of source points per
km\textsuperscript{2} within protected areas was approximately equal. In total,
we distributed 50'000 source points across protected areas. To render potential
immigrants into the study area, we placed 10'000 additional random points inside
a xx km buffer around the study area. In total, we generated 80'000 source
points, thereby resulting in 80'000 simulated dispersers, each of which moved
2'000 steps. Overall, this implied 160 Mio. simulated steps.

@@

\subsection{Network Analysis}
We coerced all simulated trajectories into a network consisting of vertices
(relocations) and edges (connections between relocations). To do so, we created
raster layers at multiple resolutions and identified each trajectory's
transition matrix on these rasters see figure xx \citep{BastilleRousseau.2018}.
We then merged the transition matrices of all trajectories and calculated
cumulative transitions between all raster-cells. This resulted in an edge-list,
containing all observed from-to connections as well as their frequency. Using
this edge-list, we generated a weighted graph using the r-package
\textit{igraph}. Based on this graph we calculated betweenness scores as well as
the degree of each raster cell. Betweenness indicates how often a specific
raster-cell lies on a shortest path between two other raster-cells and is a
useful metric to detect movement corridors. Degree, on the other hand, indicates
how many connections a raster cell has and therefore serves to illustrate the
xxx of specific nodes. When calculating betweenness, we used the transition
frequency as weighting factor. That is, a higher transition frequency
contributed to a higher betweenness score.

% Based on the simulated trajectories we identified to which other areas each
% source area is connected. For instance, if a trajectory originated at source
% point one and intersected with source areas two and three, we assumed that
% source areas two and three were within reach of source are one. To get a sense
% of the strength of connections, we also calculated how often each of these
% connections was realized. This procedure resulted in 6 x 2 edge lists which we
% further used to generate weighted networks and to calculate network metrics.

\section{Results}
\subsection{Movement Model}
Compared to the base model, the most parsimonious movement model included
several additional interactions (\Cref{MovementModel} and Table S1). The model
indicates that dispersers move directional, particularly when distant to water,
yet less so in human dominated landscapes. Furthermore, dispersers prefer large
steps, especially during main activity. In contrast, step lengths tend to be
shorter when water- or tree-cover is high. In general, dispersers avoid water,
prefer proximity to water, avoid dense tree-cover, prefer shrubs/grassland, and,
finally, avoid human dominated landscapes.
